Legacy Data Confounds Modern Studies

### 2018-04-11 Draft3.tex
introduction is functional
now have a menu for the paper

\maketitle
\tableofcontents
\clearpage
			\section{Introduction}
\subsection{Larger Sample Sizes}			
The last 5 years has seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
\lat{not a fan of this sentence}
Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
Mutations associated to a disease can be identified by comparing healthy individuals to those afflicted by a disease. 
The mutations more common in one group might be associated to the disease.
Demonstrating that these mutations are biologically relevant to the disease can be difficult.
Certain factors like age, sex or ancestral origin might be correlated to the trait in question and could potentially be mistakenly associated to a disease. 
Especially with the increasing size of cohorts, spurious associations are increasingly becoming an issue. 
For this reason, careful consideration must be taken when including individuals from different ancestral origins in these association studies. 
Benign mutations at high frequencies in one populations might be exceedingly rare in another population. 
Therefore, population wide differences in mutations must be included as covariates to avoid spurious associations.

@@

\subsection{Mutation Spectra} 
A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
The signatures from individuals from the same population will have a tendency to be more similar due to shared ancestry.
Harris et al have recently proposed that there are patterns of mutations that differ across human populations.
They argued this by showing relative differences in mutational types between and even within continental populations.

\lat{rewrite this section}
In investigating a mutational signature identified in a Japanese cohort by Harris et. al (2017), we noticed that the individuals responsible for this signal all had lower average quality scores compared to the other Japanese individuals. 
To further investigate the extent to which low quality individuals were associated to spurious mutations, we then performed a genome-wide association study (GWAS). 
We discovered \todo{000} variants that were associated to low mapping quality in \todo{000} individuals from \todo{000} of the 26 populations included in the 1000 genomes dataset. 
One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.

\subsection{Context}
This project was motivated by an unusual population genetics observation in a recent publication by Harris et. al, that the 1000 Genomes Project Japanese population seem to be partitioned in two clusters of diff mutation rates. 
This heterogeneously distributed mutational signal was unexpected as a signal of this nature could either be due to population structure, a recent mutagen, or technical bias.

Our research center had recently obtained access to a large cohort of several thousand high-quality whole genome sequences from Japanese individuals from the Fumi institute. \lat{change focus to data not center}
We started off by trying to replicate the results found in Harris et al. 2017, but after several attempts and double checking we had the right quality controls, we were unable to replicate their results with the Fumi dataset. 
Upon comparing the allele frequencies between the two datasets, we observed an unusually large number of private SNPs despite both cohorts being from the same population. 
Further investigation in the allele frequency spectra of the two Japanese datasets, indicated that the SNPs responsible for the signal were missing in the more recent dataset. 
This result raised several red flags. 
Why would a genome wide signal be concentrated in private SNPs from a legacy dataset?

Upon investigating the quality control pipelines of the more recent dataset, many of these positions uniquely found in the 1000 Genomes Project had been removed in the standard QC of the sequencing centre. 
Does this legacy dataset meet the current industry standards of quality? 
This study is the result of an investigation in the quality of the 1000 Genomes Project dataset.

After compiling all the metadata from each 1000 genomes sequencing run, we noticed a correlation between PC1 of the mutation spectrum signal and the average quality of mapped bases per individual. 
This suggested that some individuals with lower quality were driving this apparent population genetics signal. [mention fake pop structure]
Furthermore, when looking at the days when the individuals were sequenced, we noticed that individuals with this mutational signal were sequenced on the same day in the same sequencing centre.
We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.

			\section{Results}
\subsection{Average quality of mapped snps Quality Over Time}
As one would expect, the sequencing quality of individuals increases over time \ref{Figure1}. Sequencing centers improve on their laboratory techniques and sequencing technologies improve from one generation to the next. The magnitude and impact of the differences in quality within and between populations has not yet been explored.

@@

\subsection{Significant SNPs}
When we remove individuals below a quality of 30, population structure is clarified. 
The signal identified in Harris et. al is enriched in sites associated with low quality. [Need to run harris pipeline without $10^{-6}$ ]
Therefore, when these low quality sites are removed, the mutational signal is also removed. (Or in this case when we remove the individuals with qualities below 30.)

@@

Interestingly, the SNPs we identified had already been removed in higher quality datasets. 
These HQ datasets also have filtered more SNPs than the GWAS was able to identify. 
When we compare the allele frequencies between the 1000 Genomes project Japanese population to a more contemporary dataset, there are many more sites that are enriched with the same mutational signature as described above that are not identified using a GWAS. 
This suggests that we lack the power to identify all the questionable variants from these legacy datasets. 
\subsection{Imputation}
~30\% of the SNPs we identified as being associated with low quality were found to be imputed using the Michigan Imputation Server. These  should be removed from reference database

>>>>

\maketitle
\tableofcontents
\clearpage
			\section{Introduction}
	\subsection{Larger Sample Sizes}			
The last 5 years has seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
Mutations associated to a disease can be identified by comparing healthy individuals to those afflicted by a disease. 
The mutations more common in one group might be associated to the disease.
Demonstrating that these mutations are biologically relevant to the disease can be difficult.
Certain factors like age, sex or ancestral origin might be correlated to the trait in question and could potentially be mistakenly associated to a disease. 
Especially with the increasing size of cohorts, spurious associations are increasingly becoming an issue. 
For this reason, careful consideration must be taken when including individuals from different ancestral origins in these association studies. 
Benign mutations at high frequencies in one populations might be exceedingly rare in another population. 
Therefore, population wide differences in mutations must be included as covariates to avoid spurious associations.

@@

	\subsection{Mutation Spectra} 
A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
The signatures from individuals from the same population will have a tendency to be more similar due to shared ancestry.
Harris et al have recently proposed that there are patterns of mutations that differ across human populations.
They argued this by showing relative differences in mutational types between and even within continental populations.
\lat{more here about mutation spectra}

	\subsection{Context}
This project was motivated by an unusual population genetics observation in a recent publication by Harris et. al, that the 1000 Genomes Project Japanese population seem to be partitioned in two clusters of diff mutation rates. 
This heterogeneously distributed mutational signal was unexpected as a signal of this nature could either be due to population structure, a recent mutagen, or technical bias.
In investigating this mutational signature, we were unable to reproduce the results using a larger and higher quality dataset.

This study is the result of an investigation in the quality of the 1000 Genomes Project (1kGP) dataset.
To begin, we will discuss the discrepancy between the Japanese samples from the 1kGP and a more recent counterpart. 
We then consider methods to discriminate troublesome variants resulting from dated sequencing technologies. 
Furthermore, we explore more broadly how low quality variants have been propagated in the other populations included in the 1kGP.
Finally, we will discuss the impact these variants have on modern analyses.

			\section{Results}
	\subsection{A heterogeneously distributed mutational signal}			
Upon comparing the allele frequencies between the two datasets, we observed an unusually large number of private SNPs despite both cohorts being from the same population.
The SNPs responsible for the signal observed by Harris et. al (2017) were missing in the more recent dataset. 
\todo{plot SFS NAG vs JPT}
We also noticed that the individuals carrying these mutations all had lower average quality scores.
This suggested that individuals with lower quality were driving this apparent population genetics signal.
\todo{plot PC1 vs Qual}

To further investigate	 the extent to which low quality individuals were associated to spurious mutations, we then performed a genome-wide association study (GWAS) \ref{Figure2}.

We discovered \todo{000} variants that were associated to low mapping quality in \todo{000} individuals from \todo{000} of the 26 populations included in the 1000 genomes dataset. 
One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.

After compiling all the metadata from each 1000 genomes sequencing run, we noticed a correlation between PC1 of the mutation spectrum signal and the average quality of mapped bases per individual.

Furthermore, when looking at the days when the individuals were sequenced, we noticed that individuals with this mutational signal were sequenced on the same day in the same sequencing centre.
We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.

	\subsection{Average quality of mapped snps Quality Over Time}
As one would expect, the sequencing quality of individuals increases over time \ref{Figure1}. Sequencing centers improve on their laboratory techniques and sequencing technologies improve from one generation to the next. The magnitude and impact of the differences in quality within and between populations has not yet been explored.

@@

	\subsection{Significant SNPs}
When we remove individuals below a quality of 30, population structure is clarified. 
The signal identified in Harris et. al is enriched in sites associated with low quality
Therefore, when these low quality sites are removed, the mutational signal is also removed. (Or in this case when we remove the individuals with qualities below 30.)

@@

Interestingly, the SNPs we identified had already been removed in higher quality datasets. 
These HQ datasets also have filtered more SNPs than the GWAS was able to identify. 
When we compare the allele frequencies between the 1000 Genomes project Japanese population to a more contemporary dataset, there are many more sites that are enriched with the same mutational signature as described above that are not identified using a GWAS. 
This suggests that we lack the power to identify all the questionable variants from these legacy datasets.

	\subsection{Imputation}
~30\% of the SNPs we identified as being associated with low quality were found to be imputed using the Michigan Imputation Server. These  should be removed from reference database
